Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2024-01-04 manuscript/manuscript.qmd
updated intro with Betty's comments

---
title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require an extraction process.  We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

@@

A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step and have a wide range in cost. These instruments distinguish between the pigments based on their differing emission spectra. Some instruments (i.e. PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002). These instruments can be used in the field, the lab, or a combination of both.

@@

Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer. Differences in measurement methods represent a potential source of error when comparing instruments. Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison. In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (@liu_chlorophyll_2021). Estimation problems can also result from differing community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022). Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM. Some instruments offer a yellow substance correction to help account for CDOM while others do not (@cremella_interpretation_2018). How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types. The goal of this paper is to compare different handhelds or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.

@@

Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction. Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in-vivo module. chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath. Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer. Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).

@@

The results from this study indicate that in-vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll. Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements. Instruments measuring phycocyanin were better able to handle differences in the phytoplankton community between these study ponds that cause chlorophyll measurements to be less strongly correlated with extracted results. Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.

@@

This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies. Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in-vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in-vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.

>>>>

---
title: "Comparing fluorometric methods ( in vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require advanced laboratory processes (e.g. pigment extraction).  We compared measurements from five in vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and  in vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between  in vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from  in vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in vivo fluorometry and phycocyanin for cyanoHAB monitoring efforts and when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

@@

A variety of instruments have been developed to measure chlorophyll *a* and/or phycocyanin using in vivo fluorescence without an extraction step. These instruments distinguish between the pigments based on their differing emission spectra. Some instruments (i.e. PhycoProbe) use multiple leds at varying wavelengths to cause fluorescence of different algal types, thus creating a finer scale breakdown of community composition (@beutler_fluorometric_2002). These instruments can be used in the field, the lab, a combination of both, and have a wide range in cost.

@@

Field instruments may be placed directly in a waterbody or require sample water to be poured into a cuvette and then placed into a handheld fluorometer. Differences in measurement methods represent a potential source of error when comparing instruments. Additionally, instruments may report values in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison. In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light (@liu_chlorophyll_2021). Estimation problems can also result from differing phytoplankton/cyanobacteria community composition and growth stage, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior (@rousso_chlorophyll_2022). Additionally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM. Some instruments offer a yellow substance correction to help account for CDOM while others do not (@cremella_interpretation_2018).

Overall, the physical properties and phytoplankton community composition of a waterbody can cause instruments to estimate pigment concentrations differently. How the different instruments handle these various potential situations can affect their accuracy and precision, and the ability to compare across instrument types. The goal of this paper is to compare several handheld or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.

@@

Within 24 hours of collection, field samples will be analyzed in triplicate using the PhycoProbe, CyanoFluor, and the Trilogy (in-vivo), as well as filtered onto 0.7 µm pre-ashed glass fiber filters and frozen at -20 °C for solvent-based extraction. Sample water filtered through a 0.22 µm syringe filter was used to correct for yellow substances on the PhycoProbe, CyanoFluor, and Trilogy in vivo module. chlorophyll *a* extraction using 90% acetone followed after a 20 minute period in a sonicating water bath. Phycocyanin extraction followed a 15 minute period in a sonicating water bath and used a 50 mM phosphate buffer. Extracted samples were analyzed using a Turner Trilogy with a chla na module (chlorophyll) and an orange module (phycocyanin).

@@

The results from this study indicate that in vivo phycocyanin is better related to extracted phycocyanin when compared to the same measurements for chlorophyll. Similarly, phycocyanin from all instruments was better correlated to cell counts than chlorophyll measurements. Instruments measuring phycocyanin were better able to handle differences in the phytoplankton community between these study ponds that cause chlorophyll measurements to be less strongly correlated with extracted results. Both extracted phycocyanin and chlorophyll were strongly correlated with cell counts across the six study ponds suggesting that pigments are indeed a good proxy for cell counts at least for some waterbodies.

@@

This study supports the use of extracted chlorophyll and phycocyanin as an alternative to cell counts for the study waterbodies. Currently, in vivo chlorophyll *a* is typically used for routine monitoring of cyanoHABs, but this study found that phycocyanin to be better correlated with extracted phycocyanin across a variety of instruments. Therefore, measuring in vivo phycocyanin may be a good proxy for cell counts. There are a variety of instruments capable of measuring phycocyanin using in vivo fluorometry across a spectrum of prices. We hope this study will encourage further research and consideration of using in vivo phycocyanin as alternative to cell counts and in situations were extractive techniques are not able to be used.
